\section{Other Related Work}\label{sec:related}
We survey related work not discussed previously.
% \textsection\ref{sec:intro}-\ref{sec:logic} and \textsection\ref{sec:outro}.

A memory consistency model for a shared-memory multiprocessor defines the
values that a read may return.  For a survey of hardware models, see
\citep{AlglaveThesis}. For software models, see
\citep{DBLP:journals/toplas/Lochbihler13,DBLP:phd/ethos/Batty15}.  For an
attempt to bridge the two, see \citep{DBLP:journals/pacmpl/PodkopaevLV19}.
\citet{DBLP:conf/pldi/PultePKLH19} present an operational model of
\armeight{} in the style of \cite{DBLP:conf/popl/KangHLVD17}.

In our previous work~\cite{2019-sp}, we introduced the notion of pomsets with
preconditions.  In \citeyear{2019-sp}, we studied \emph{micro\hyp{}architecture},
where failed speculative execution is visible via cache effects.  Here we
presented an \emph{architectural} model, which allows us to impose
\emph{consistency}, ignoring failed speculative execution, and \emph{causal
  strengthening}.  In \citeyear{2019-sp}, we used 3-valued pomsets, as
opposed to the simple pomsets used here; see \textsection\ref{sec:limits} for
further discussion.  The previous paper was not focused on memory models, and
thus did not prove the soundness of compiler optimizations, the absence of
thin air reads, efficient implementability, or \drfsc{}.

% Both \cite{2019-sp} and this paper are inspired by \citeauthor{Batty17}'s
% [\citeyear{Batty17}] call for a compositional approach to the semantics of
% relaxed memory concurrency.  
% in relaxed 
% outlines the program of trying to
% understand relaxed dependency modularly, inspiring \citet{2019-sp} and
% \citet{DBLP:conf/esop/PaviottiCPWOB20}.  % ; impressively, the latter underlies a
% recent proposal to update the ISO-C11 standard.
%(http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2019/p1780r0.html).

Our model shares important structural elements with that of
\citet{DBLP:conf/esop/PaviottiCPWOB20}, who provide a fix for the thin-air
problem in C11.  Like us, they use true concurrency semantics to identify
(in)dependence in an execution and thus calculate the preserved program order
explicitly.  Our definition of parallel composition allows events to
coalesce, taking preconditions via disjunction---this is mirrored by
\citeauthor{DBLP:conf/esop/PaviottiCPWOB20}'s definition of coproduct.  We
only compose downclosed pomsets---this is mirrored by their condition on
$\leq^{\mathcal{X}}$ during coproduct (discussed in their \textsection6.3).
Nonetheless, the papers have different goals, leading to different outcomes.
For example, their model compiles efficiently to non-\mca{} architectures; our
model clearly does not!  Conversely, our model provides an intrinsic
characterization of optimizations, such as redundant read elimination, which
only hold in their model up to observational refinement.
% In contrast to our LDRF
% theorem, they prove an internal version of DRF, ie that race free programs
% have SC behavior, without necessarily settling the question of discovering
% races by SC executions.

% \citeauthor{DBLP:conf/esop/PaviottiCPWOB20} develop a comprehensive array of
% tools to deal with the testing and validation of weak memory models.  In
% future work, we hope to adapt this suite to our setting.



True concurrency techniques have been applied to relaxed memory by
\citet{DBLP:conf/esop/CenciarelliKS07}, \citet{Castellan},
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
\citet{DBLP:conf/cgo/ChakrabortyV17}.  See
\cite[\textsection8]{DBLP:journals/lmcs/JeffreyR19} for a discussion.
A partial order approach to weak memory was sketched by \citet{brookes} and
fleshed out for TSO by \citet{DBLP:journals/lmcs/KavanaghB19}.  Their action
labels include buffers, encoding the operational behavior of TSO inside the
pomsets themselves.

There is a rich literature on the use of transformations over SC
executions to model relaxed memory:
%
\citet{Saraswat:2007:TMM:1229428.1229469}
aimed to describe a \jmm-like model this way.  \drfsc\ holds, but \citet{SP} 
discovered that it permits \oota\ behavior.
%
\citet{DBLP:conf/popl/DemangeLZJPV13} developed \textsc{bmm}, which permits
reordering of a relaxed write with a following relaxed read.
% The paper develops an axiomatic semantics and studies effects on compilation.  
\textsc{bmm} is designed as a restriction of the \jmm\ that compiles efficiently to
\tso.  It requires fencing on other architectures.
%
\citet{DBLP:conf/fm/LahavV16} characterized \tso\ as being derived by
considering Write-Read (\textsf{WR}) reordering and Read-After-Write
(\textsf{RAW}) elimination.  They also showed that the release acquires of
C11 are less expressive than considering \textsf{WR} and \textsf{RAW}
together with thread-inlining.  Our paper is inspired by their implicit
challenge: ``Some memory models can be defined via transformations.  But
there is more to weak memory than transformations.''

% \paragraph{Modeling the Invalidation of Load Buffering.} 

% In this paragraph, we argue that our paper is relevant even to the reader
% who desires a semantics that invalidates ``load buffering,'' or ``reads from the future.''
% There are two related lines of research that suggest to prohibit the
% reordering of loads with following stores.
% \begin{itemize}
% \item \citet{Dolan:2018:BDR:3192366.3192421} imposed this restriction in
%   order to ``bound data races in space and time.''
%   Consider the following program:
%   % In the following program $r,s$ are registers, and $x,y,z,u,v,w$
%   % are shared variables.
%   % \begin{align*}
%   %   & (u \GETS 0 \SEMI y \GETS 1 \SEMI v\REL  \GETS 1 \SEMI z \GETS 2 \SEMI r \GETS u \SEMI y \GETS 3) \\[-.5ex]
%   %   \PAR & (y \GETS 2 \SEMI  \IF{v\ACQ} \THEN w \GETS 1 + y -y \SEMI u \GETS 1 \SEMI x \GETS 1  \FI \SEMI s \GETS u \SEMI  y \GETS 2)
%   % \end{align*}
% \begin{displaymatharrayrl}
%   &x\GETS1\SEMI y\GETS1\SEMI v\REL\GETS1\SEMI z\GETS1\SEMI
%   \IF{w}\THEN x\GETS2\FI
%   \\\PAR
%   &y\GETS2\SEMI \aReg{}\GETS v\ACQ\SEMI z\GETS2\,\SEMI
%   \IF{\aReg{}}\THEN w\GETS{y-y+x}\FI
% \end{displaymatharrayrl}
% \begin{displaymath}
%     \begin{tikzpicture}[node distance=1.2em and .8em,baselinecenter]
%     \event{a1}{\DW{x}{1}}{}
%     \event{a2}{\DW{y}{1}}{right=of a1}
%     \event{a3}{\DWRel{v}{1}}{right=of a2}
%     \event{a4}{\DW{z}{1}}{right=of a3}
%     \pox{a1}{a2}
%     \pox{a2}{a3}
%     \pox{a3}{a4}
%     \event{b1}{\DW{y}{2}}{below=of a2}
%     \event{b2}{\DRAcq{v}{1}}{below=of a3}
%     \event{b3}{\DW{z}{2}}{right=of b2}
%     %\co{b1}[right]{a2}
%     \event{b4}{\DR{y}{?}}{right=of b3}
%     \event{b5}{\DR{y}{?}}{right=of b4}
%     %\co{b3}[right]{a4}
%     \event{b6}{\DR{x}{1}}{right=of b5}
%     \pox{b1}{b2}
%     \pox{b2}{b3}
%     \pox{b3}{b4}
%     \pox{b4}{b5}
%     \pox{b5}{b6}
%     \rf{a3}{b2}
%     \event{b7}{\DW{w}{1}}{right=of b6}
%     \event{a5}{\DR{w}{1}}{above=of b7}
%     \event{a6}{\DW{x}{2}}{right=of a5}
%     \pox{a4}{a5}
%     \pox{a5}{a6}
%     \pox{b6}{b7}
%     \rf{b7}{a5}
%     \rf[out=-18,in=162]{a1}{b6}
%     %\race{b7}{a5}
%     %\race{a6}{b6}
%   \end{tikzpicture}
% \end{displaymath}
% In their model, one can deduce that $w$ must be assigned $1$, despite the
% past race on $y$, concurrent race on $z$ and future race on $x$.
% % In their model, one can reason sequentially about the  assignment to $w$
% % and deduce that $x=1$ implies $z= 1$ despite the concurrent races on $u$, and
% % the past and future races on $y$, where past and future are stated with
% % respect to the assignment to $w$.

% \item In the C11 family of models, eg. see
%   \citep{Boehm:2014:OGA:2618128.2618134,DBLP:conf/pldi/LahavVKHD17,DBLP:conf/oopsla/VafeiadisN13},
%   load buffering is forbidden to avoid \oota. The elegant discussion in
%   \citet{BoehmOOTA} elucidates the issues via a series of ``\oota'' like
%   examples.  In particular, to rule out example~\ref{rfub}.
% \end{itemize}

% Our model can easily capture models that forbid load buffering.  It
% necessitates two changes.
% \begin{itemize}
% \item Modify the semantic rule for read to remove the possibility of an
%   internal read:
%   \begin{align*}
%     \sem{\aReg\GETS\REF{\cExp}\SEMI \aCmd} & =
%     \textstyle\bigcup_{\aLoc=\REF{\bVal}}\; ({\cExp}=\bVal) \guard \textstyle\bigcup_\aVal\; (\DR\aLoc\aVal) \prefix \sem{\aCmd}[\aLoc/\aReg] 
%   \end{align*}
% \item Modify item \ref{pre-read}b in the definition of prefixing to require
%   order from a read to subsequent write.  Item \ref{pre-read}b$'$ becomes: if
%   $\aEv$ is a write then $\cEv\lt'\aEv$.  In \ref{pre-read}b$'$ we removed
%   the ``or  $\labelingForm'(\aEv)$ implies $\labelingForm(\aEv)$.''
% \end{itemize}

% While \citet{Dolan:2018:BDR:3192366.3192421} already describes DRF and the
% compilation to \armeight, our approach yields the following new insights.
% \begin{itemize}
% \item Compositional treatment of temporal invariants
%   from \textsection\ref{sec:logic} holds mutatis mutandis, since all
%   executions of the restricted model are already accounted for in the general
%   model.

% \item With regards to single-threaded optimizations in
%   \textsection\ref{sec:opt}, our approach provides different methods to prove
%   optimizations, which complement the extant methods of
%   \citet{Dolan:2018:BDR:3192366.3192421} with data sensitivity; for example,
%   our methods provides a simple proof of the transformation of
%   $\IF{\aExp}\THEN x \GETS 1 \SEMI \aCmd \ELSE x\GETS 1 \SEMI \bCmd\FI$ to
%   $x \GETS 1 \SEMI \IF{\aExp}\THEN \aCmd \ELSE \bCmd\FI$.

% \end{itemize}

%\paragraph{Relationship to Models of Speculation.}

\section{Limitations}
\label{sec:limits}

% Our \drfsc{} theorem does not address fences or \RMW{}s.  \emph{On the upside,
% we have proven a \emph{local} \drfsc{} theorem, which is stronger
% than the standard \drfsc{} result.}
%\todo{assume fixed sized access}

 Our work has several limitations, each of which provides an opportunity for
future research.  % In several cases, these limitations have allowed us to
% \emph{surpass prior work in significant ways}.

We have not modeled loops or functions.  Loops and recursive functions
introduce complexities\allowbreak{}---such as liveness and continuity---that
are orthogonal to the main topic of the paper.

We have not modeled general sequencing of the form $(\aCmd\SEMI\aCmd')$.  The
definition of prefixing (\textsection\ref{sec:pop}) is relatively simple
since we only prepend one action at a time.
% We also do not include types,
% allocation, garbage collection, etc.  % This is the norm for
% work on relaxed memory: The problem is already hard
% enough. % \emph{On the upside, we do allow
  % thread creation via $\!\!\PAR\!\!$, which is unusual.}

We have not modeled mixed-size access
\cite{DBLP:conf/popl/FlurSPNMGSBS17,DBLP:conf/pldi/WattPPBDFPG20}.  
\armeight{} captures multibyte access using multiple events related by
\emph{same instruction} $(\rsi)$ \cite{alglave-git}. To capture the use of
$\rsi$ when defining \emph{locally ordered before} $(\rlob)$, it is likely
sufficient to modify \ref{5b}.  To capture the use of $\rsi$ when defining
\emph{observation} $(\rob)$, it is likely sufficient to modify \ref{rf3} and
\ref{rf4}.
%The atomicity of mixed size access is not as strong as that of \RMWs.

We have not attempted validate all program transformations that involve
synchronization, fences or \RMWs{}
\cite{DBLP:conf/popl/VafeiadisBCMN15,DBLP:phd/hal/Morisset17}.  Nonetheless,
our semantics validates roach motel (\ref{AcqW}, \ref{RelW}), redundant load
(\ref{RL}), fence removal (\ref{RF}), and store forwarding (\ref{SF}).  We
expect that dead store elimination (\ref{DS}) generalizes to non-relaxed
access by generalizing $\relfilt[]{\aLoc}$ in Definition \ref{def:cover}.
Some transformations are not sound in the model, but we expect them to be
provable as metaproperties.  For example, access-mode strengthening (such as
replacing $\modeRLX$ by $\modeRA$) is valid up to an equivalence that ignores
access modes on actions.  Other transformations worth studying include
commuting adjacent synchronizations and implementing non-relaxed access using
relaxed access and fences.  Lock elision and access-mode weakening are also
interesting, although these are only sound in certain contexts.  We expect
all of these transformations to be valid, given an appropriate notional of
validity.
\begin{comment}
https://preshing.com/20131125/acquire-and-release-fences-dont-work-the-way-youd-expect/

Cannot encode R/A actions with actions+fences...

A release operation prevents preceding memory operations from being delayed
past it (a;Rel =/=> Rel;a)
 
A release fence prevents preceding memory operations from being delayed past
subsequent writes (a;FR;w =/=> w;a;FR)

An acquire operation prevents subsequent memory operations from being advanced
before it (Acq;a =/=> a;Acq)

An acquire fence prevents subsequent memory operations from being advanced
before prior reads (r;FA;a =/=> FA;a;r)

https://www.modernescpp.com/index.php/fences-as-memory-barriers

StoreLoad: Full fence allows a store before to be reordered with respect to a
load after (wx;F;ry) ===> (ry;F;wx)

StoreLoad+LoadLoad: Release fence also allows (rx;FR;ry) ===> (ry;FR;rx)

StoreLoad+StoreStore: Acquire fence also allows (wx;FR;wy) ===> (wy;FR;wx)

LoadStore: No fence allows a prior load to reorder w.r.t. a subsequent store
(rx;FR;wy) =/=> (wy;FR;rx)

https://preshing.com/20120710/memory-barriers-are-like-source-control-operations/
Good news is that a fullFence does it.

Bizarrely, it seems this is not supported in C++... You have to go to assembly.
\end{comment}


\begin{comment}
  \color{red} perhaps add example here from referee 1, or discuss \ref{Internal2}.

Although our model forces some ordering within the thread that includes the
RA access, threads that lack RA access are free to reorder.  Perhaps this
gives the desired effect.  For example
$$
(y:=1; x^{ra}:=1) || (s:=x; z:=1)
$$
allows the execution $(Wz1) \leq (Wy1) \leq (W^{ra}x1) \leq (Rx1)$.  
(Here, we are using augmentation to add order from $Wz1$ to $Wy1$.)

We do expect that lock-based implementation of atomics should be sound, if
locks were introduced as first class entities, but we have not explored this.
Intuitively, lock elision can be validated by using roach motel to move the
entire program into the scope of the lock.
\end{comment}
 % Accommodating
% such optimizations will likely require changes to the definition of
% $\textsf{cover}$ (page \pageref{def:cover}) and $\textsf{read}$ (page
% \pageref{page:readsat}).
% \emph{On the upside, 
  % Corollary~\ref{seqcompleteness} is the first abstract characterization of
  % valid transformations for relaxed access.}



\labeltext{The}{page:logic:limits} logic we presented in \textsection\ref{sec:logic} is only strong enough
to prove a few examples.  \citet{DBLP:conf/esop/SvendsenPDLV18} presented a
different a logic, capable of showing that the following program cannot write
2:
\begin{align*}
  \taglabel{OOTA5}
    ( y\GETS x+1
    \PAR
    x\GETS y ) && \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{2}}{right=of rx}
        \po{rx}{wy}
        \event{ry}{\DR{y}{1}}{right=3em of wy}
        \event{wx}{\DW{x}{1}}{right=of ry}
        \po{ry}{wx}
        % \rf{wy}{ry}
        \rf[out=170,in=10]{wx}{rx}
      \end{tikzinline}}
\end{align*}
The attempted execution is \emph{not} allowed by our semantics, since there is no write
to fulfill $(\DR{y}{1})$.  % Proving this requires the ability to reason about
% values.
% \labeltext{The}{page:logic:limits} logic we presented in \textsection\ref{sec:logic} is only strong enough
% to prove a few examples.
%
As another example,
consider the following, from \citet[Fig.~3]{DBLP:journals/pacmpl/ChakrabortyV19}:
\begin{gather*}
  \taglabel{OOTA6}
  \begin{gathered}
  x\GETS 2\SEMI
  \IF{x\NOTEQ2}\THEN y\GETS 1 \FI
  \PAR
  x\GETS 1\SEMI
  r\GETS x\SEMI
  \IF{y}\THEN x\GETS 3 \FI
  \\\nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{wx2}{\DW{x}{2}}{}
  \event{rx3}{\DR{x}{3}}{right=of wx2}
  \event{wy1}{\DW{y}{1}}{right=of rx3}
  \po{rx3}{wy1}
  \event{wx1}{\DW{x}{1}}{right=2em of wy1}
  \event{rx2}{\DR{x}{2}}{right=of wx1}
  \wk{wx1}{rx2}
  \event{ry1}{\DR{y}{1}}{right=of rx2}
  \event{wx3}{\DW{x}{3}}{right=of ry1}
  \po{ry1}{wx3}
  \wk[in=165,out=15]{rx2}{wx3}
  \rf[in=-170,out=-10]{wy1}{ry1}
  \rf[in=170,out=10]{wx2}{rx2}
  \rf[out=-170,in=-10]{wx3}{rx3}
  \wk[out=-170,in=-10]{wx1}{wx2}
  \wk{wx2}{rx3}
    \end{tikzinline}}
\end{gathered}
\end{gather*}
The attempted execution is \emph{not} allowed by our semantics, due to the
evident cycle.  Intuitively, it is not possible for the left thread to
read $3$ for $x$ when the right thread reads $2$.  Proving this may require a
logic with modalities to deal with intervening writes and coherence.  Surprisingly, this outcome is allowed by the promising
semantics \cite{DBLP:conf/popl/KangHLVD17}.
\citeauthor{DBLP:journals/pacmpl/ChakrabortyV19} developed \weakestmo{} to
address this example; however, \weakestmo{} does not address \ref{OOTA4}.  
% \emph{On the upside, our logic is powerful enough to show that our semantics
%   disallows \oota{} executions of \ref{OOTA4} --- \weakestmo{} and the
%   promising semantics both permit \oota{} executions of \ref{OOTA4} (see \textsection\ref{sec:promising}).}

% Our results have not been formally verified, but only because we do not have
% the resources to do it ourselves.  It is worth noting that the current
% emphasis on formal verification has the unintended effect of making research
% more conservative: Large libraries are built, and large institutions with
% teams of graduate students build them.

Our model realizes \emph{multi-copy atomicity} (\mca).  Thus it will not
compile efficiently to non-\mca{} architectures, such as \ppc\ and \armseven.
To do so, one cannot include the order required by \ref{rf4} in pomset order.
In \cite{2019-sp}, we modeled programs using \emph{3-valued pomsets}, using
the \emph{weak} relation ($\gtNsp$) for \ref{rf4} and \ref{5b}.  This does
\emph{not} provide a suitable model for non-\mca{} behavior, however, since
it disallows \ref{MCA2}\footnote{Following \cite{DBLP:journals/dc/Lamport86},
  3-valued pomset require: (1) if $d\leq e$ then $d\gtNsp e$, (2) if
  $d\leq e$ and $e\gtNsp d$ then $d=e$, and (3) if $c\leq d \gtNsp e$ or
  $c\gtNsp d\leq e$ then $c\gtNsp e$.  \ref{MCA2} is disallowed by (2).
  Top-level 3-valued pomsets also require that $\gtNsp$ is a partial order
  per-location; a sufficient condition is (4) if $d\gtNsp e$ and $e\gtNsp d$
  then $d$ and $e$ do not touch the same location.  Although \ref{MCA1} is
  allowed, its two-thread variant is not, due to the combination of
  \emph{semi\hyp{}transitivity (3)} and \emph{partial coherence (4)}.}.



In the discussion of \ref{MCA2}, we noted that our model does not enforce
order between reads dues to address and control dependencies.  This has
implications for Java's final field semantics.  Consider:
\begin{gather*}
  \taglabel{addr}
  \begin{gathered}
  (r\GETS 1 \SEMI \REF{r}\GETS 0 \SEMI \REF{r}\GETS 1\SEMI  x^\modeRA\GETS r)
  \PAR
  (r\GETS x^\modeRA \SEMI s \GETS \REF{r})
  \\[-1ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DW{\REF{1}}{0}}{}
      \event{a2}{\DW{\REF{1}}{1}}{right=of a1}
      \wk{a1}{a2}
      \event{a4}{\DWRel{x}{1}}{right=of a2}
      \sync{a2}{a4}
      \event{b1}{\DR[\modeRA]{x}{1}}{right=3em of a4}
      \event{b2}{\DR{\REF{1}}{0}}{right=of b1}
      \sync{b1}{b2}
      \rf{a4}{b1}
      \wk[out=170,in=10]{b2}{a2}
    \end{tikzinline}}
  \end{gathered}
\end{gather*}
If we changed $x^\modeRA$ to $x^\modeRLX$, then there would be no order from
$(\DR{x}{1})$ to $(\DR{\REF{1}}{0})$, and the execution would be allowed.  In
order to allow this relaxation in certain cases without invalidating the
publication of the ``field initializer'' $(\DW{\REF{1}}{1})$, it may be
desirable to distinguish address dependencies from other dependencies---doing
so would likely require two separate preconditions for each event.
% .  without
% distinguishing address dependency from other things, or keeping thread ids in
% actions.  ARM requires that we do not order writes that only have a control
% dependency or that are fulfilled by an external writes.

% In the concurrent there is an inherent tension between ``simplicity'' and
% compiler optimizations.  Where these are in tension, we have opted to prefer
% the compiler.  In particular, we do not enforce address and control
% dependencies between reads.  This can cause counterintuitive outcomes, such
% as the following.

% Final fields example without address dependency:
% r=new C; r.f=1; Fence1; x=r   ||   s1=x; Fence2; if (s1) {s2=s1.f}
% r=100; [r]=1; Fence1; x=r     ||   s1=x; Fence2; if (s1) {s2=[s1]}
% without Fence2 we allow the outcome s1=100, s2=0

% \section{Conclusions}
% \label{sec:outro}

% We have defined a relaxed memory model, using standard tools from mathematics,
% and demonstrated that it satisfies the fundamental criterion for
% software memory models, namely ``the memory model must strike a balance
% between ease-of-use for programmers and implementation flexibility for system
% designers.''
% We have made two substantial contributions:

% In \textsection\ref{sec:logic}, we proposed a cyclical proof rule for
% parallel composition of temporal properties in the style of
% \citet{Abadi:1993:CS:151646.151649}.  This provides an objectively
% falsifiable alternative to the notion of \emph{out of thin air} (\oota)
% execution, which has been famously difficult to prove or disprove
% \cite{DBLP:conf/esop/BattyMNPS15,BoehmOOTA}.

% In \textsection\ref{sec:opt}, we formalized a connection between Hoare
% logic and the sequential fragment of our language, allowing us to make a
% general statement about transformations.  Thus, we present a solution to a
% problem that has been open since \citet{DBLP:conf/esop/CenciarelliKS07}
% discovered reordering of independent statements is invalid in the JMM
% \cite{Manson:2005:JMM:1047659.1040336}.

% These contributions rest on a semantics that supports a simple, executable
% prescription of \emph{essential dependencies}, calculated using logic and
% manifest in pomset order.  This provides a
% clear interface between the language user and the compiler writer, and a
% clear bridge from the compiler writer to computer architecture.

% We presented a denotational semantics using pomsets with preconditions.  The
% model may also be expressible in the operational framework of
% \citet{DBLP:conf/esop/FerreiraFS10}, which parameterizes a standard SC
% operational semantics with respect to a set of program transformations.




\endinput

\section{Conclusions}
We have defined a relaxed memory model, using standard tools from the
semantics and demonstrated that it satisfies the fundamental criterion for
software memory models, namely ``the memory model must strike a balance
between ease-of-use for programmers and implementation flexibility for system
designers.''

One key lesson we draw is that any software relaxed memory model should
support the cyclical composition proof rule of
\citet{Abadi:1993:CS:151646.151649} for temporal properties.  We posit that all
the benefits of \oota, such as typing guarantees, can be derived from this
principle. 

We conclude with a suggestion for the consideration of programming language
designers.  The semantics of a programming language should support a clear
and executable prescription of dependency in a program.  Such a specification
provides a nice interface between the language user and the compiler writer,
and a clear bridge from the compiler writer to architectures.

Our current paper is to be viewed as a step towards this goal.  We will
explore developing an executable specification based on these ideas in future
work.
 


\endinput
\section{SCRAPS}

Some \oota\ examples can be proven if we extend the logic with value
quantification---we leave the details to future work.  Consider
\cite[(\textsc{rng})]{DBLP:conf/esop/SvendsenPDLV18}:
\begin{gather*}
  % \VAR x\GETS0\SEMI \VAR y\GETS0\SEMI \VAR z\GETS0\SEMI %\\[-.5ex]
  \begin{aligned}
    ( y\GETS x+1
    \PAR
    x\GETS y ) && \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{2}}{right=of rx}
        \po{rx}{wy}
        \event{ry}{\DR{y}{1}}{right=3em of wy}
        \event{wx}{\DW{x}{1}}{right=of ry}
        \po{ry}{wx}
        % \rf{wy}{ry}
        \rf[out=170,in=10]{wx}{rx}
      \end{tikzinline}}
  \end{aligned}
  \\
  \hbox{$
    \begin{gathered}
      [\forall u,v.\;(\once\DW{y}{u} \land \once\DW{y}{v}) \Rightarrow ((u \EQ v) \vee (u \EQ 0) \vee (v \EQ 0))]
      \\[-.5ex]
      \land [\DW{x}{1} \Rightarrow \once\DR{y}{1}]
      \land [\DW{y}{2} \Rightarrow \once\DR{x}{1}]
      % \land [\DW{z}{2} \Rightarrow \once\DR{y}{2}]
    \end{gathered}$}
\end{gather*}
% \begin{displaymath}
%   \VAR x\GETS0\SEMI \VAR y\GETS0\SEMI \VAR z\GETS0\SEMI  %\\[-.5ex]
%   (
%     y\GETS x+1\SEMI z \GETS y
%   \PAR
%     x\GETS y
%     %   )
% \end{displaymath}
% \begin{scope}
%   \small
%   \begin{gather*}
%     [(\once\DW{y}{v_1} \land \once\DW{y}{v_2}) \Rightarrow ((v_1 = v_2) \vee (v_1 = 0) \vee (v_2 = 0))] \land\\[-.5ex]
%     [\once\DW{x}{1} \Rightarrow \once\DR{y}{1}] \land [\once\DW{y}{2}
%     \Rightarrow \once\DR{x}{1}] \land [\once\DW{z}{2} \Rightarrow
%     \once\DR{y}{2}]
%   \end{gather*}
% \end{scope}
The attempted execution is disallowed since there is no write to fulfill
$\DR{y}{1}$.  Closing $x$ and $y$ and using the last two conjuncts in the
invariant above:
% last conjunct, we have
% \begin{math}
%   % \once\DW{z}{2} \Rightarrow \once\DR{y}{2}
%   % \once\DW{z}{2} \Rightarrow \once\DW{y}{2}
%   \DW{y}{2} \Rightarrow \once\DR{x}{1}.
% \end{math}
% Closing $x$, we then have
\begin{math}
  % \once\DW{z}{2} \Rightarrow \once\DW{x}{1}
  \DW{y}{2} \Rightarrow \once\DW{y}{1}.
\end{math}
From the first conjunct, we then derive
\begin{math}
  \DW{y}{2} \Rightarrow \FALSE,
\end{math}
as required.

Other \oota\ examples may require that we extend the logic to deal with
intervening writes or coherence. The following outcome is allowed by the
promising semantics \cite{DBLP:conf/popl/KangHLVD17}, but not in \weakestmo{}
\cite[Fig.~3]{DBLP:journals/pacmpl/ChakrabortyV19} nor in our semantics, due
to the cycle:
\begin{gather*}
  \label{weakestmo}
  x\GETS 2\SEMI
  \IF{x\NOTEQ2}\THEN y\GETS 1 \FI
  \PAR
  x\GETS 1\SEMI
  r\GETS x\SEMI
  \IF{y}\THEN x\GETS 3 \FI
  \\\nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
  \event{wx2}{\DW{x}{2}}{}
  \event{rx3}{\DR{x}{3}}{right=of wx2}
  \event{wy1}{\DW{y}{1}}{right=of rx3}
  \po{rx3}{wy1}
  \event{wx1}{\DW{x}{1}}{right=2em of wy1}
  \event{rx2}{\DR{x}{2}}{right=of wx1}
  \wk{wx1}{rx2}
  \event{ry1}{\DR{y}{1}}{right=of rx2}
  \event{wx3}{\DW{x}{3}}{right=of ry1}
  \po{ry1}{wx3}
  \wk[in=165,out=15]{rx2}{wx3}
  \rf[in=-170,out=-10]{wy1}{ry1}
  \rf[in=170,out=10]{wx2}{rx2}
  \rf[out=-170,in=-10]{wx3}{rx3}
  \wk[out=-170,in=-10]{wx1}{wx2}
  \wk{wx2}{rx3}
    \end{tikzinline}}
\end{gather*}


\subsection{RFUB}
Let $\aCmd$ be the right thread in \ref{rfub}.
\begin{align*}
  \aCmd = r\GETS y\SEMI \IF{r \NOTEQ 1} \THEN z\GETS 1\SEMI r\GETS 1\FI  \SEMI x\GETS r 
\end{align*}
Then, $\hoare{\TRUE}{\aCmd}{x=1} $ and $\hoare{y \neq 1}{\bCmd}{z=1}$.  Thus, the traditional sequential semantics {\em does not} attribute any cause for the write of $1$ to $x$, as seen by the execution:
\begin{tikzdisplay}[node distance=1.5em]
 \event{ry}{\DR{y}{1}}{}
 \event{wx}{\DW{x}{1}}{right=of ry}
 \event{wa}{\DW{z}{1}}{below=of rx}
 \rf{ry}{wa}
\end{tikzdisplay}
Consider the
following, where all locations are initialized to $0$.  In \textsection\ref{sec:logic} we provide machinery to prove that \oota{} is
incapable of writing $1$.  The question is whether this should be possible
for \rfub, which changes \oota{} only to include a \emph{Register assignment
  From an Unexecuted Branch} \cite{BoehmOOTA}:?  
\begin{align*}
  \label{oota}  \tag{\textsc{oota}}
  y\GETS x
  \PAR&
  r\GETS y\SEMI
  x\GETS r
  \\
  \tag{\rfub1}
  \label{rfub}
  y\GETS x
  \PAR&
  r\GETS y\SEMI
  \IF{r \NOTEQ 1} \THEN z\GETS 1\SEMI r\GETS 1\FI  \SEMI x\GETS r 
\end{align*}
%and the following \emph{Out Of Thin Air} litmus test:
The change in the second
thread from \oota{} to \rfub{} is not a valid refinement under Hoare logic:
\rfub{} validates the triple $\hoare{\TRUE}{\aCmd}{x=1}$, but \oota{} does
not.  As a result, it is expected that \rfub{} may have additional behaviors.

The program writes $1$ to $x$ in both branches of the conditional.  Further,
the writes to $z$ and $x$ in the then-branch of the conditional are
independent.  Therefore, it is sensible for a compiler to hoist the write to
$x$ out of the conditional.  In our semantics, there is no dependency from
the read of $y$ to the write to $x$
The execution in question is:
\begin{displaymathsmall}
  \begin{tikzcenter}[node distance=1.5em]
    \event{a1}{\DR{x}{1}}{}
    \event{a2}{\DW{y}{1}}{right=of a1}
    \po{a1}{a2}
  \end{tikzcenter}
  \Bigm\|
  (\DR{y}{1})
  \prefix
  \left(
    \begin{tikzcenter}[node distance=.5em]
      \event{b1}{r\NOTEQ1\mid\DW{z}{1}}{}
      \event{b2}{\DW{x}{1}}{right=of b1}
    \end{tikzcenter}
    % \begin{tikzcenter}[node distance=.5em]
    %   \event{b1}{r\NOTEQ1\mid\DW{z}{1}}{}
    %   \event{b2}{r\NOTEQ1\mid\DW{x}{1}}{right=of b1}
    % \end{tikzcenter}
    % \biggm\|
    % \begin{tikzcenter}[node distance=1.5em]
    %   \event{c1}{r\EQ1\mid\DW{x}{1}}{}
    % \end{tikzcenter}
  \right)
\end{displaymathsmall}
With an internal read of $y$, we have:
\begin{displaymathsmall}
  \begin{tikzcenter}[node distance=1.5em]
    \event{a1}{\DR{x}{1}}{}
    \event{a2}{\DW{y}{1}}{right=of a1}
    \po{a1}{a2}
  \end{tikzcenter}
  \Bigm\|
    \begin{tikzcenter}[node distance=1.5em]
      \event{b1}{y\NOTEQ1\mid\DW{z}{1}}{}
      \event{b2}{\DW{x}{1}}{right=of b1}
    \end{tikzcenter}
\end{displaymathsmall}
but the precondition $y\NOTEQ1$ cannot be satisfied locally.
With an external read of $(\DR{y}{1})$, we can discharge the precondition, but in this
case, the predicate becomes $1\NOTEQ 1$. 
\begin{displaymathsmall}
  \begin{tikzcenter}[node distance=1.5em]
    \event{a1}{\DR{x}{1}}{}
    \event{a2}{\DW{y}{1}}{right=of a1}
    \po{a1}{a2}
  \end{tikzcenter}
  \Bigm\|  
    \begin{tikzcenter}[node distance=1.5em]
      \nonevent{b1}{1\NOTEQ1\mid\DW{z}{1}}{}
      \event{b2}{\DW{x}{1}}{right=of b1}
      \event{b0}{\DR{y}{1}}{left=of b1}
      \po{b0}{b1}
    \end{tikzcenter}
\end{displaymathsmall}


