\section{Data Race Free Behaviors are Sequentially Consistent}
\label{sec:sc}

% For any $\aPSS$, then $\closed(\aPSS)$ is set enriched with useless reads
% (preserving augmentation closure) and where we remove any event whose
% precondition is not a tautology.
% \begin{definition}
%   Let $\addRead(\aPSS)$ be the set $\aPSS'$ where $\aPS'\in\aPSS'$ whenever
%   there is $\aPS\in\aPSS$ such that:
%   $\Event' = \Event\cup{\cEv}$,
%   ${\le'} \supseteq {\le}$, 
%   ${\gtN'} \supseteq {\gtN}$,
%   and
%   $\labelingAct'(\cEv) = (\DR{\aLoc}{\aVal})$ and $\labelingAct'(\aEv) = \labelingAct(\aEv)$,
% \end{definition}
% Then $\fclosed(\aPSS)$ 



In this section, we prove the SC-DRF theorem, which states that any program
that lacks data races under the SC semantics must only have executions that
are compatible with SC executions.  We present the result for programs of the
form $\vec{\aLoc}\GETS\vec{0}\SEMI\aCmd$, where $\aCmd$ is restriction-free.  Thus, all memory locations are initialized to $0$, and internal actions only arise from (intra-thread) implicit reads.

We say that two actions have a \emph{data-race conflict} if at least one
action is a write and the other is a write, read, or internal read to the
same location.  Define the relation $\reco$ so that $(\aEv,\bEv)\in{\reco}$
if $\aEv\gtN\bEv$ and $\aEv$ and $\bEv$ have a data-race conflict.

The program $x\GETS1\PAR x\GETS2$ is considered to have an SC data race, but
$x\GETS1\SEMI x\GETS2$ does not.  In our semantics the only difference
between these is that $x\GETS1\SEMI x\GETS2$ enforces weak order between the
writes.  Note also that the
$\sem{x\GETS1\SEMI a\GETS y}=\sem{a\GETS y\SEMI x\GETS1}$, yet these two must
be distinguished in SC, as per the load-buffering and store-buffering litmus tests.

In order to define SC executions and SC data races, it is necessary to
augment our semantics to record program order and internal reads.  We extend the definitions in
\textsection\ref{sec:semantics} with
${\rpox}\subseteq{\Event}\times{\Event}$, defined as follows:
\begin{itemize}
\item
  ${\rpox'} = {\rpox}$
  when $\aPSS'=\aPSS\aSub$
  or $\aPSS'=\aForm\guard\aPSS$
\item
  ${\rpox'} = {\rpox}\restrict{\Event'}$
  when $\aPSS'=\nu\aLoc\st\aPSS$
\item
  ${\rpox'} = {\rpox}^1\cup{\rpox}^2$
  when $\aPSS'=\aPSS^1\parallel\aPSS^2$
\item
  ${\rpox'} = {\rpox}\cup\{(\cEv,\aEv)\mid\aEv\in\Event\}$
  when $\aPSS'=\aAct\prefix\aPSS$ and $\Event' = \Event \cup \{\cEv\}$
\end{itemize}

Define the relation ${\rrfx}$ so that $(\aEv,\bEv)\in{\rrfx}$ if $\aEv$
writes $\aLoc$, $\bEv$ reads $\aLoc$, and for any $\cEv$ that writes $\aLoc$
either $\cEv\gtN\aEv$ or $\bEv\gtN\cEv$.  Let $\IDAcq$ be the identity
relation on acquire events, and likewise $\IDRel$ on release events.  Now
define $\rsw$ and $\rhb$\footnote{For simplicity, the definition of $\rsw$
  does not include release sequences or fences.  For example, we consider
  \begin{math}
    (\DWRel{x}{1})\prefix(\DW{x}{2}) \parallel (\DRAcq{x}{2})
  \end{math}
  to be racy, whereas this pomset is race-free using release sequences.  If
  we include release sequences and fences, then $\rhb$ relates more events
  and thus there are fewer races. Our results hold under either definition.}.
\begin{align*}
  {\rsw} &= \IDRel; ({\rrfx}\setminus{\rpox}); \IDAcq
  \\
  {\rhb} &= ({\rpox} \cup {\rsw})^+
\end{align*}
Note that our semantics guarantees that ${\rsw}\subseteq{\lt}$.

A pomset has a \emph{data race} if there are events $\aEv$ and $\bEv$ such
that
\begin{itemize}
\item $\aEv$ and $\bEv$ are unordered by $\rhb$,
\item $\labelingForm(\aEv)$ and $\labelingForm(\bEv)$ are tautologies, and
\item $\labelingAct(\aEv)$ and $\labelingAct(\bEv)$ have a data-race conflict.
\end{itemize}

%The semantics of programs includes SC executions.
\begin{definition}
  Let $\semsc{\aCmd}$ be the subset of $\sem{\aCmd}$ such that
  $\aPS\in\semsc{\aCmd}$ whenever $\aPS$ is a top-level pomset and ${\lt}\cup{\rpox}$ is acyclic.
% ${\lt}\cup{\rpox} \cup {\reco}$ is acyclic.
\end{definition}
%\begin{itemize}
%\item ${\le}\cup{\rpox} \cup {\reco}$ is acyclic ,
% \item prefixing ($\prefix$) and composition ($\parallel$) take disjoint union, and
%\item all reads are denoted by explicit actions.
%\end{itemize}
We argue that this definition is sufficient to capture sequential
consistency: Any total order that linearizes the acyclic relation is
consistent with strong order ($\lt$) and the program order ($\rpox$).  Since $\le$ contains $\reco$, the consistency with $\reco$ ensures that only the last write to a location is
read in such a total order.

%\begin{remark}\label{generator}
% In the rest of this section, we consider ``top-level'' programs of the form
% $\aCmd = \VAR\vec{\aLoc}\SEMI
%     \vec{\aLoc}\GETS\vec{0}\SEMI
%     \vec{\bLoc}\GETS\vec{0}\SEMI
%     \FENCE\SEMI
%     (\aCmd_1 \PAR \cdots \PAR \aCmd_n)
% $.
% We will consider closed and complete executions of $\aCmd$. We use $\semClosed{\aCmd}$ to stand for the subset of $\sem{\aCmd}$  with only  the pomsets that are $\vec{\aLoc}, \vec{\bLoc}$ closed.  
We only consider \emph{generators}, which are top-level pomsets that are
minimal with respect to augmentation and implication.  Since we are
considering finite programs without loops, the pomsets in the semantics of
threads are finite.  Thus, there are no infinite descending chains of
augmentations.


%\end{remark}

We prove the following theorem in \textsection\ref{drfproof}.
\begin{theorem}
  Let $\aPS$ be a generator for $\aCmd$.
  \begin{itemize}
  \item If $\aPS$ does not have a data race, $\aPS \in \semsc{\aCmd}$.
  \item If $\aPS$ has a data race, then there exists
    $\aPS'\in \semsc{\aCmd}$ that has a data race.
  \end{itemize}
\end{theorem}
A key step of this proof is an analysis of the closure properties of the semantics.  In order to perform this fine grained analysis of dependency, we describe a variant of the semantics using three valued pomsets defined below.  
\begin{definition}
  \label{def:tvalpom}
  A \emph{\tvalpom} is a tuple
  $(\Event, {\sle}, {\gtN},
  \labeling)$, such that
  \begin{itemize}
   \item $(\Event, {\gtN},
  \labeling)$ is a pomset, and 
\item ${\sle} \subseteq {\gtN}$ is a partial order. 
  \end{itemize}
\end{definition}
\citet{DBLP:conf/esop/HuthJS01} call $\slt$ a ``must transition''
and $\geN$ a ``may transition''--- we have used  the terms ``strong order'' and ``weak order'' respectively.  In the pictures so far, we have anticipated this development and adopted
\citeauthor{DBLP:journals/dc/Lamport86}'s notation, drawing $\slt$ as a solid arrow ``$\xpo$'' and $\geN$ dashed ``$\xwk$''.  The intuitive temporal meaning of $ \aEv \lt \bEv$ is that $\aEv$ {\em must} strictly precede $\bEv$, whereas $ \aEv \gtN \bEv$ is intended to connote that $\aEv$ may precede $\bEv$. Thus, if  $ \neg (\aEv \gtN \bEv)$, $\aEv$ cannot precede $\bEv$.  
 
The semantics of programs in the three-valued model proceed mutatis mutandis, with details described in the appendix.  This finer analysis of necessary and possible dependency allows us to establish the existence of pomsets in the semantics as we search for sequential witnesses to data races.    We provide some illustrative examples below.

\noindent
For the program:
\begin{align*}
\mbox{Thread 1: } &\VAR y\GETS 0 \SEMI \aReg \GETS y  \SEMI x \GETS 1  \SEMI \\[-.5ex]
\mbox{Thread 2: } &\VAR x\GETS 0 \SEMI \bReg \GETS x \SEMI y \GETS 1  \SEMI 
\end{align*}
\begin{displaymathsmall}
\mbox{From    }
\begin{tikzcenter}[node distance=1em]
\event{wy0}{\DW{y}{0}}{}
\event{ry1}{\DR{y}{1}}{right=of wy0}
\event{wx1}{\DW{x}{1}}{right=of ry1}
\event{wx0}{\DW{x}{0}}{below=of wy0}
\event{rx1}{\DR{x}{1}}{right=of wx0}
\event{wy1}{\DW{y}{1}}{right=of rx1}
\rf{wx1}{rx1}
\rf{wy1}{ry1}
\wk{wx0}{rx1}
\wk{wy0}{ry1}
\end{tikzcenter}
\mbox{  infer  }
\begin{tikzcenter}node distance=1em]
\event{wy0}{\DW{y}{0}}{}
\event{ry1}{\DR{y}{0}}{right=of wy0}
\event{wx1}{\DW{x}{1}}{right=of ry1}
\event{wx0}{\DW{x}{0}}{below=of wy0}
\event{rx1}{\DR{x}{0}}{right=of wx0}
\event{wy1}{\DW{y}{1}}{right=of rx1}
\rf{wx0}{rx1}
\rf{wy0}{ry1}
\wk{rx1}{wx1}
\wk{ry1}{wy1}
\end{tikzcenter}
\end{displaymathsmall}
For the program:
\begin{align*}
\mbox{Thread 1: } &\VAR y\GETS 0 \SEMI   x \GETS 1  \SEMI \aReg \GETS y  \SEMI \\[-.5ex]
\mbox{Thread 2: } &\VAR x\GETS 0 \SEMI  y \GETS 1  \SEMI  \bReg \GETS x \SEMI
\end{align*}
\begin{displaymathsmall}
\mbox{From  }
\begin{tikzcenter}[node distance=1em]
\event{wy0}{\DW{y}{0}}{}
\event{wx1}{\DW{x}{1}}{right=of wy0}
\event{ry0}{\DR{y}{0}}{right=of wx1}
\event{wx0}{\DW{x}{0}}{below=of wy0}
\event{wy1}{\DW{y}{1}}{right=of wx0}
\event{rx0}{\DR{x}{0}}{right=of wy1}
\rf[bend right]{wx0}{rx0}
\rf[bend left]{wy0}{ry0}
\wk{rx0}{wx1}
\wk{ry0}{wy1}
\wk{wx0}{wx1}
\wk{wy0}{wy1}
\end{tikzcenter}
\ \mbox{ infer }
\begin{tikzcenter}[node distance=1em]
\event{wy0}{\DW{y}{0}}{}
\event{wx1}{\DW{x}{1}}{right=of wy0}
\event{ry0}{\DR{y}{1}}{right=of wx1}
\event{wx0}{\DW{x}{0}}{below=of wy0}
\event{wy1}{\DW{y}{1}}{right=of wx0}
\event{rx0}{\DR{x}{1}}{right=of wy1}
\rf{wx1}{rx0}
\rf{wy1}{ry0}
\wk{wx0}{wx1}
\wk{wy0}{wy1}
\end{tikzcenter}
\end{displaymathsmall}
For the program:
\begin{align*}
\mbox{Thread 1: } &\VAR x\GETS 1 \\[-.5ex]
\mbox{Thread 2: } &\VAR x\GETS 0 
\end{align*}
\begin{displaymathsmall}
\mbox{From  }
\begin{tikzcenter}[node distance=1em]
\event{wy0}{\DW{x}{1}}{}
\event{wx0}{\DW{x}{0}}{right=of wx0}
\wk{wy0}{wx0}
\end{tikzcenter}
\mbox{ infer   }
\begin{tikzcenter}[node distance=1em]
\event{wy0}{\DW{x}{1}}{}
\event{wx0}{\DW{x}{0}}{right=of wx0}
\wk{wx0}{wy0}
\end{tikzcenter}
\end{displaymathsmall}


\endinput

We say that $\aCmd$ has an \emph{SC race} if there is some pomset in $\semsc{\aCmd}$ that contains a data race.


In this section we show that if $\semsc{\aCmd}\subseteq\sem{\aCmd}$ has only
race-free executions, then each pomset $\aPS\in\sem{\aCmd}$ is ``equivalent''
to some $\aPS'\in\semsc{\aCmd}$, where $\aPS'$ may have more events, but
preserves labeling and has less order.

We say that $\aPS\suborder\aPS'$ if there is an injection
$\inj:\Event'\rightarrow\Event$ such that:
\begin{itemize}
\item $\labelingAct'(\aEv) = \labelingAct(\inj(\aEv))$
\item $\labelingForm'(\aEv)$ implies $\labelingForm(\inj(\aEv))$
\item $\labelingForm(\bEv)$ implies $\bigvee_{\aEv\in\inj^{-1}(\bEv)}(\labelingForm'(\aEv))$
\item $\aEv\le'\bEv$ implies $\inj(\aEv)\le\inj(\bEv)$
\item $\aEv\gtN'\bEv$ implies $\inj(\aEv)\gtN\inj(\bEv)$
\end{itemize}

\begin{theorem}
  If $\semsc{\aCmd}$ contains only race-free executions and
  $\aPS\in\sem{\aCmd}$ then there exists some $\aPS'\in\semsc{\aCmd}$ such
  that $\aPS\suborder\aPS'$.
\end{theorem}
% \begin{proof}
%   \begin{itemize}
%   \item
%     \begin{math}
%       \sem{\SKIP}
%       =
%       \{ \emptyset \} 
%     \end{math}
%   \item
%     \begin{math}
%       \sem{\FENCE_{\aF}\SEMI \aCmd}
%       =
%       (\DF{\aF}) \prefix \sem{\aCmd}
%     \end{math}
%   \item
%     \begin{math}
%       \sem{\aLoc\GETS\aExp\SEMI \aCmd}
%       =
%       \textstyle\bigcup_\aVal\; \bigl((\aExp=\aVal) \guard (\DW\aLoc\aVal) \prefix \sem{\aCmd}\bigr)[\aExp/\aLoc]
%     \end{math}
%   % \item
%   %   \begin{math}
%   %     \sem{\REL\aLoc\GETS\aExp\SEMI \aCmd}
%   %     =
%   %     \textstyle\bigcup_\aVal\; \bigl((\aExp=\aVal) \guard (\DWRel\aLoc\aVal) \prefix \sem{\aCmd}\bigr)[\aExp/\aLoc]
%   %   \end{math}
%   \item
%     \begin{math}
%       \sem{\aReg\GETS\aLoc\SEMI \aCmd}
%       =
%       \sem{\aCmd}[\aLoc/\aReg] \cup \textstyle\bigcup_\aVal\; (\DR\aLoc\aVal) \prefix \sem{\aCmd}[\aLoc/\aReg]
%     \end{math}
%   % \item
%   %   \begin{math}
%   %     \sem{\ACQ\aReg\GETS\aLoc\SEMI \aCmd}
%   %     =
%   %     \textstyle\bigcup_\aVal\; (\DRAcq\aLoc\aVal) \prefix \sem{\aCmd}[\aLoc/\aReg]
%   %   \end{math}
%   \item
%     \begin{math}
%       \sem{\IF{\aExp} \THEN \aCmd \ELSE \bCmd \FI}
%       =
%       \bigl((\aExp \neq 0) \guard \sem{\aCmd}\bigr) \parallel \bigl((\aExp=0) \guard \sem{\bCmd}\bigr)
%     \end{math}
%   \item
%     \begin{math}
%       \sem{\aCmd \PAR \bCmd}
%       =
%       \sem{\aCmd} \parallel \sem{\bCmd}
%     \end{math}
%   \item
%     \begin{math}
%       \sem{\VAR\aLoc\SEMI \aCmd}
%       =
%       \nu \aLoc \st \sem{\aCmd}
%     \end{math}
% \end{itemize}
  
% \end{proof}
% \end{theorem}


% To define compatibility, we extend the definitions of
% \textsection\ref{sec:semantics} to include an additional order: $\rird$.
% \begin{itemize}
% \item
%   ${\rird'} = {\rird}$
%   when $\aPSS'=\aPSS\aSub$
%   or $\aPSS'=\aForm\guard\aPSS$
% \item
%   ${\rird'} = {\rird}\restrict{\Event'}$
%   when $\aPSS'=\nu\aLoc\st\aPSS$
% \item
%   ${\rird'} = {\rird}^1\cup{\rird}^2$
%   when $\aPSS'=\aPSS^1\parallel\aPSS^2$
% \item
%   ${\rird'} = {\rird}\cup\{(\cEv,\aEv)\mid\labelingForm(\aEv) \;\text{is dependent on}\; \aLoc\}$
%   when $\aPSS'=\aAct\prefix\aPSS$, $\aAct$ writes $\aLoc$, and $\Event' = \Event \cup \{\cEv\}$
% \end{itemize}

% From $\rird$, we define ${\rrb}={\rird}^{-1};{\gtN}$.

% We want that if there is an execution:
% \begin{tikzdisplay}[node distance=1em]
%   \event{a}{\DW{\aLoc}{1}}{}
%   \event{b}{\DW{\bLoc}{1}}{below right=1em and 6em of a}
%   \event{c}{\DW{\aLoc}{2}}{above right=1em and 1em of b}
%   \wk{a}{c}
%   \ird{a}{b}
%   \rb{b}{c}
% \end{tikzdisplay}
% Then there is also
% \begin{tikzdisplay}[node distance=1em]
%   \event{a}{\DW{\aLoc}{1}}{}
%   \event{b}{\DW{\bLoc}{1}}{below right=1em and 6em of a}
%   \event{c}{\DW{\aLoc}{2}}{above right=1em and 1em of b}
%   \event{r}{\DR{\aLoc}{1}}{below right=.1em and 2em of a} 
%   \wk{a}{c}
%   \rf{a}{r}
%   \po{r}{b}
% \end{tikzdisplay}


% To see that we need $[\aExp/\aLoc]$ in the rule for write, rather than $[\aVal/\aLoc]$
% consider example:
% \begin{verbatim}
% r=y; if (r) {x=r} else {x=r}; s=x; if (r==s) {z=1}
% \end{verbatim}
% or simplified:
% \begin{verbatim}
% r=y;x=r;s=x; if(s==r){z=1}
% \end{verbatim}
% If you read 37 for $y$, then the predicate on \texttt{Wz1} before the
% read is either $r=r$ or $v=r$, where $v=37$, for example.  In one case you
% get a dependency and in the other you do not.

