\section{Fences, Read-Modify-Write, and Address Calculation}
\label{sec:variants}

We extend the model to include additional features: fences, read-modify-write
actions (\RMWs), and address calculation. The proofs given later in the paper
extend to include these features.

\myparagraph{Fences}

Syntactic fences
``$\FENCE^{\fmode} \SEMI \aCmd$'' have corresponding  actions: $(\DFS{\fmode})$.  The \emph{syntactic fence mode}
$(\fmode \!\!\BNFDEF\!\! \mREL \!\BNFSEP\! \mACQ \!\BNFSEP\! \mSC)$
is either \emph{release}, \emph{acquire}, or \emph{sequentially-consistent}.

$(\DFS{\mREL})$ is a release. $(\DFS{\mACQ})$ is an acquire.
$(\DFS{\mSC})$ is both a release and an acquire.
\begin{align*}
  \sem{\FENCE^{\fmode}\SEMI \aCmd} & \eqdef
  (\DFS{\fmode}) \prefix \sem{\aCmd}
\end{align*}
With no further changes, syntactic fences would impose exactly the same order
as synchronization actions.  This is sufficient to simulate $\mSC$
accesses, since $\mSC$ fences are very strong.  However, it is
insufficient to simulate $\mRA$ accesses.
Thus we add the following requirements to Definition \ref{def:prefix}:%  of
\begin{enumerate}
\item[{\labeltextsc[P5e]{(P5e)}{5e}}] if $\bEv$ reads, and $\aEv$ is an acquiring fence, then
  $\bEv \le' \aEv$, and
\item[{\labeltextsc[P5f]{(P5f)}{5f}}] if $\bEv$ is a releasing fence, and $\aEv$ writes, then
  $\bEv \le' \aEv$.
\end{enumerate}
Consider the following variant of \ref{Pub1}:
\begin{gather*}
  \taglabel{Pub2}
  \begin{gathered}
    x\GETS0\SEMI %y\GETS0\SEMI
    x\GETS 1\SEMI \FENCE^\mREL\SEMI y \GETS1
    \PAR
    r\GETS y \SEMI \FENCE^\mACQ\SEMI s\GETS x
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{fr}{\DFS{\mREL}}{right=of wx1}
        \event{wy1}{\DW{y}{1}}{right=of fr}
        \event{ry1}{\DR{y}{1}}{right=2.5em of wy1}
        \event{fa}{\DFS{\mACQ}}{right=of ry1}
        \event{rx0}{\DR{x}{0}}{right=of fa}
        \sync{wx1}{fr}
        \sync{fr}{wy1}
        \sync{ry1}{fa}
        \sync{fa}{rx0}
        \rf{wy1}{ry1}
        \wk{wx0}{wx1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
\ref{5f} requires that $(\DFS{\mREL})\le(\DW{y}{1})$.  \ref{5e} requires
that $(\DR{y}{1})\le(\DFS{\mACQ})$.  The other order involving fences is
required by \ref{5c}.  The attempted execution is
\emph{invalid}: the stale read $(\DR{x}{0})$ violates \ref{rf4}.

As for \ref{RL}, redundant fence elimination \eqref{RF} follows from \ref{1},
regardless of the access mode.
\begin{align*}
  \taglabel{RF}
  \sem{\FENCE^\fmode \SEMI \FENCE^\fmode\SEMI\aCmd} &\supseteq 
  \sem{\FENCE^\fmode \SEMI \aCmd}
\end{align*}

Our semantics does not suffer from overly weak fencing (\ref{SC3}) or a lack
of fence cumulativity (\ref{SC4}).  The following examples, from
\citet[Figs.~5 and 6]{DBLP:conf/pldi/LahavVKHD17}, are allowed by both the
original C11 and the model of \citet{DBLP:conf/popl/BattyDW16}.  We omit
$0$\hyp{}initialization in these examples:
\begin{gather*}
  \taglabel{SC3}
  \begin{gathered}
    x\GETS1
    \PAR
    r\GETS x\SEMI   
    \FENCE^{\mSC}\SEMI
    r\GETS y  
    \PAR
    y\GETS 1 \SEMI
    \FENCE^{\mSC}\SEMI
    r\GETS x  
    \\[-.1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a1}{\DW{x}{1}}{}
        \event{b1}{\DR{x}{1}}{right=3em of a1}
        \event{b2}{\DFS{\mSC}}{right=of b1}
        \sync{b1}{b2}
        \event{b3}{\DR{y}{0}}{right=of b2}
        \sync{b2}{b3}
        \event{c1}{\DW{y}{1}}{right=3em of b3}
        \event{c2}{\DFS{\mSC}}{right=of c1}
        \sync{c1}{c2}
        \event{c3}{\DR{x}{0}}{right=of c2}
        \sync{c2}{c3}
        \wk{b3}{c1}
        \rf{a1}{b1}
        \wk[out=-170,in=-10]{c3}{a1}
      \end{tikzinline}}
  \end{gathered}
  \\
  \taglabel{SC4}
  \begin{gathered}
    x\GETS1\SEMI   
    z^\mRA\GETS1\SEMI   
    \PAR
    r^\mRA\GETS z\SEMI   
    \FENCE^{\mSC}\SEMI
    r\GETS y  
    \PAR
    y\GETS 1 \SEMI
    \FENCE^{\mSC}\SEMI
    r\GETS x  
    \\[-.1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a1}{\DW{x}{1}}{}
        \event{a2}{\DWRel{z}{1}}{right=of a1}
        \sync{a1}{a2}
        \event{b1}{\DRAcq{z}{1}}{right=2em of a2}
        \event{b2}{\DFS{\mSC}}{right=of b1}
        \sync{b1}{b2}
        \event{b3}{\DR{y}{0}}{right=of b2}
        \sync{b2}{b3}
        \event{c1}{\DW{y}{1}}{right=2em of b3}
        \event{c2}{\DFS{\mSC}}{right=of c1}
        \sync{c1}{c2}
        \event{c3}{\DR{x}{0}}{right=of c2}
        \sync{c2}{c3}
        \wk{b3}{c1}
        \rf{a2}{b1}
        \wk[out=-170,in=-10]{c3}{a1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
The executions are disallowed, due to the evident cycles.  While these
results are immediate in our model, it is worth noting that they are anything
but immediate in the various models of C11.

\myparagraph{Read-Modify-Write} We discuss \RMW\ operations that work on a
single location in memory, such as \emph{fetch-and-add} ($\FADD$) and
\emph{compare-and-swap} ($\CAS$).  These operations can be modeled using read/write
actions or using an additional relation between events.  The second approach
is more general and less obvious, therefore we explain it here.

In Definition \ref{def:mmpomset}, we require that a \emph{(memory model) pomset}
be a tuple $(\Event, {\le}, \labeling, \rrmw)$, where ${\rrmw}\subseteq{\le}$
relates the two events of a successful \RMW.  Additionally, we require that:
\begin{itemize}
\item If $\cEv$, $\aEv$ write the same $x$, $\cEv\gtN \aEv$ and $\bEv \xrmw \aEv$ then  $\cEv\gtN \bEv$.
\item If $\cEv$, $\aEv$ write the same $x$, $\bEv\gtN \cEv$ and $\bEv \xrmw \aEv$ then  $\aEv\gtN \cEv$.
\end{itemize}

The constituent events of an \RMW{} may coalesce with other events and other
\RMW{}s, but we prevent coalescing of separate \RMW{}s in the semantics of
$\!\PAR\!$ and prefixing.  We elide the obvious and tedious semantic rules that
generate $\rrmw$.

This definition ensures atomicity, disallowing executions such as
\cite[Ex.~3.2]{DBLP:journals/pacmpl/PodkopaevLV19}:
\begin{gather*}
  \taglabel{RMW1}
  \begin{gathered}
    \aLoc\GETS0\SEMI\bReg\GETS \FADD^{\mRLX,\mRLX}(\aLoc,1)
    \PAR
    x\GETS 2\SEMI s\GETS x
    \\
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a2}{\DR{x}{0}}{}
        \event{a1}{\DW{x}{0}}{left=of a2}
        \rf{a1}{a2}
        \event{a3}{\DW{x}{2}}{right=of a2}
        \wk{a2}{a3}
        \event{b2}{\DW{x}{1}}{right=of a3}
        \event{b3}{\DR{x}{1}}{right=of b2}
        \rmw[out=-15,in=-165]{a2}[below]{b2}
        \wk{a3}{b2}
        \rf{b2}{b3}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}

By using two actions rather than one, the definition allows examples such as the
following, which is allowed by \armeight{} 
\cite[Ex.~3.10]{DBLP:journals/pacmpl/PodkopaevLV19}:
\begin{gather*}
  \taglabel{RMW2}
  \begin{gathered}
    r \GETS y\SEMI
    z \GETS r
    \PAR
    r\GETS z\SEMI
    x\GETS 0\SEMI
    s\GETS \FADD^{\mRLX,\mRA}(x,1) \SEMI
    y\GETS s{+}1
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a1}{\DR{y}{1}}{}
        \event{a2}{\DW{z}{1}}{right=of a1}
        \po{a1}{a2}
        \event{b1}{\DR{z}{1}}{right=3em of a2}
        \rf{a2}{b1}
        \event{b2}{\DW{x}{0}}{right=of b1}
        \event{b3}{\DR{x}{0}}{right=of b2}
        \rf{b2}{b3}
        \event{b4}{\DWRel{x}{1}}{right=2em of b3}
        \rmw{b3}{b4}
        \event{b5}{\DW{y}{1}}{right=of b4}
        \sync[out=-15,in=-165]{b1}{b4}
        \po[out=-20,in=-160]{b3}{b5}
        \rf[out=170,in=10]{b5}{a1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}

\myparagraph{Address Calculation}

In the definition of a data model, we require
that locations have the form $\aLoc\!\!\BNFDEF\!\!\REF{\cVal}$, where $\cVal$
is a value.  Expressions may include neither memory locations nor the
operator $\REF{\cExp}^{\amode}$.
In our example language, we update the syntax of commands:
\begin{gather*}
  \aCmd
  \BNFDEF \cdots
  \BNFSEP \aReg\GETS\REF{\cExp}^{\amode}\SEMI \aCmd 
  \BNFSEP \REF{\cExp}^{\amode}\GETS\aExp\SEMI \aCmd
\end{gather*}
Address calculation can be encoded using the conditional.  We give the
semantics simply by expanding this encoding.  Applying this technique to Candidate
\ref{cand:ord}, we arrive at the following:
\begin{align*}
  \sem{\aReg\GETS\REF{\cExp}^\amode\SEMI \aCmd} & \eqdef
  \;\textstyle\parallel_\cVal (\cExp=\cVal) \guard \bigl(
  \textstyle\bigcup_\aVal\;
  (\DRmode{\REF{\cVal}}\aVal) \prefix \sem{\aCmd} [\REF{\cVal}]/\aReg]
  \bigr)
  \\
  \sem{\REF{\cExp}^\amode\GETS\aExp\SEMI \aCmd} & \eqdef
  \;\textstyle\parallel_\cVal (\cExp=\cVal) \guard 
  \bigl(\textstyle\bigcup_\aVal\; (\aExp=\aVal \mid \DWmode{\REF{\cVal}}\aVal) \prefix \sem{\aCmd}[\aExp/\REF{\cVal}]\bigr)
\end{align*}
The same technique to can be applied to Definition \ref{def:cover}---we
elide the lengthy but obvious definition. For degenerate programs that include only
constant references (every expression $\REF{\cExp}^\amode$ satisfies
$\cExp=\cVal$, for some $\cVal$), the resulting definition produces exactly
the same executions as before.

The rewrites listed in \textsection\ref{sec:props}-\ref{sec:refine} remain valid, with the
following generalization: For address expressions $\REF{\aExp}$ and
$\REF{\bExp}$, replace $\aLoc=\bLoc$ by provable equality of $\aExp$ and
$\bExp$, and $\aLoc\neq\bLoc$ by provable inequality.  % Operations on sets can

In Definition \ref{def:cover}, we were able to ensure disjunction closure by
performing targeted, local weakening via $\Rdis{\aLoc}{\aVal}$.  This is much
more difficult to do with address calculation, since a single bit of syntax
can refer to multiple locations.  Consider that
\begin{math}
  \sem{[r] \GETS 0\SEMI [0]\GETS \BANG r}
\end{math}
includes both of the following pomsets (``$\BANG$'' is logical
negation---``$\BANG \aExp$'' evaluates to $1$ if
$\aExp$ is $0$, and $0$ otherwise):
\begin{align*}
  \hbox{\begin{tikzinline}[node distance=2em]
      \eventl{d}{a}{r\EQ0\mathbin{\mid}\DW{[0]}{0}}{}
      \eventl{e}{b}{r\EQ0\mathbin{\mid}\DW{[0]}{1}}{right=of a}
      \wk[out=-15,in=-165]{a}{b}
    \end{tikzinline}}%\;\;\cdots
  &&
  \hbox{\begin{tikzinline}[node distance=2em]
      \eventl{c}{a}{r\EQ1\mathbin{\mid}\DW{[1]}{0}}{}
      \eventl{d}{b}{r\EQ1\mathbin{\mid}\DW{[0]}{0}}{right=of a}
    \end{tikzinline}}%\;\;\cdots
\end{align*}
Thus, the disjunction closure also includes both of the following: % By using \!$\PAR$\!, it also includes:
\begin{align*}
  \hbox{\begin{tikzinline}[node distance=2em]
      \eventl{d}{a}{r\EQ0\lor r\EQ1\mathbin{\mid}\DW{[0]}{0}}{}
      \eventl{e}{b}{r\EQ0\mathbin{\mid}\DW{[0]}{1}}{right=of a}
      \wk[out=-15,in=-165]{a}{b}
    \end{tikzinline}}%\;\;\cdots
  &&
  \hbox{\begin{tikzinline}[node distance=2em]
      \eventl{c}{a}{r\EQ1\mathbin{\mid}\DW{[1]}{0}}{}
      \eventl{d}{b}{r\EQ0\lor r\EQ1\mathbin{\mid}\DW{[0]}{0}}{right=of a}
    \end{tikzinline}}%\;\;\cdots
\end{align*}
In this example, the $d$ events that coalesce come from inconsistent executions.
This is possible because the $d$ events originate from different commands.

\labeltext{Because}{page:strengthening} we do not enforce order between
reads, there is some danger that address calculations could introduce
anomalous behaviors that arise \emph{out of thin air} (\oota).  Consider the following program, where
initially $x=0$, $y=0$, $\REF{0}=0$, $\REF{1}=2$, and $\REF{2}=1$.  It should
only be possible to read $0$, disallowing the attempted execution below:
\begin{gather*}
  \label{strengthening}\tag{\textsc{addr1}}
  \begin{gathered}
    \aReg\GETS y\SEMI \bReg\GETS \REF{\aReg}\SEMI x\GETS \bReg
    \PAR
    \aReg\GETS x\SEMI \bReg\GETS \REF{\aReg}\SEMI y\GETS \bReg
    \\
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a1}{\DR{y}{2}}{}
        \event{a2}{\DR{\REF{2}}{1}}{right=of a1}
        \event{a3}{\DW{x}{1}}{right=of a2}
        \po{a2}{a3}
        \po[out=10,in=170]{a1}{a3}
        \event{b1}{\DR{x}{1}}{right=3em of a3}
        \event{b2}{\DR{\REF{1}}{2}}{right=of b1}
        \event{b3}{\DW{y}{2}}{right=of b2}
        \po{b2}{b3}
        \po[out=10,in=170]{b1}{b3}
        \rf[out=-170,in=-10]{b3}{a1}
        \rf{a3}{b1}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
Although no order is enforced between reads, the read-to-write order induced
by the semantics is sufficient to prohibit this \oota{} behavior.  Note the
intermediate state of $\sem{\bReg\GETS \REF{\aReg}\SEMI x\GETS \bReg}$:
\begin{gather*}
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a2}{\aReg{=}2\mid\DR{\REF{2}}{1}}{}
      \event{a3}{\aReg{=}2\land 1{=}1\mid\DW{x}{1}}{right=of a2}
      \po{a2}{a3}
    \end{tikzinline}}
\end{gather*}
The precondition $(\aReg{=}2\mid\DW{x}{1})$ is required by causal strengthening
(Def.~\ref{def:mmpomset}). % : if $\bEv\le\aEv$ then $\labelingForm(\aEv)$

